### 2018-09-28 paper/paper.tex
Update introduction

Because of the time-consuming nature of experimental determination of RNA structures, a considerable amount of work has been put into computational (as opposed to experimental) approaches. For secondary structure prediction, comparative analysis is the gold standard  \cite{Gutell1992-hu}. But accurate comparative analysis typically requires a large number of homologous sequences, which may or may not be available. On the other hand, the prevailing {\em computational} approaches  XXX References??? XXX are based on an important and controversial assumption. Namely, that the structures of  non-coding RNA molecules, {\em in vivo}, are in thermal equilibrium. If this were the case, and if the configuration energies could be accurately specified, then secondary (and in principle, even tertiary) structures could be explored through Monte Carlo sampling of the Gibbs equilibrium distribution. (XXX or is it the Boltzmann equilibrium??? XXX) Alternatively, depending on the nature of the energy landscape, secondary structures could be approximated by identifying the minimum free energy (MFE) configuration  \cite{Mathews1999-hc,Zuker1999-rc}. However, even if we put aside the necessity for approximations, the biological relevance of equilibrium configurations has been a source of misgivings at least since 1969, when Levinthal pointed out that the time required to equilibrate might be too long by many orders of magnitude  \cite{Levinthal1969-hi}. In light of these observations, and considering the ``frustrated'' nature of the folding landscape, many have argued that when it comes to structure prediction for macromolecules, kinetic accessibility is more relevant than equilibrium thermodynamics  \cite{Higgs2000-gj,Flamm2008-ir,Baker1994-px}.

@@

We will introduce quantitative measures of ambiguity and demonstrate their statistical relationships to native secondary structure and to qualitative distinctions between RNA families.
Using a somewhat arbitrary convention, we will
refer to sequences of length four (four consecutive nucleotides) as {\em segments}. Keeping this convention in mind, each line of statistical
evidence is based on the same formal definition of the ``local ambiguity'' of a given segment. This is simply the number of its complementary pairs in the molecule. 
When we refer to the {\em location} of a segment, we will mean the location of the first element of the segment, counting from 5$^\prime$ to 3$^\prime$, and  
when we refer to the local ambiguity of a location, we will mean the local ambiguity of the segment at that location. 
Local ambiguity is an intrinsic property of the primary structure. We will be interested in exposing its statistical relationships to any given candidate secondary structure of the same molecule. For this purpose, for any particular secondary structure we distinguish three different kinds of locations:
\begin{description}
	\item[\textit{Single}:] Locations where all nucleotides in the corresponding segment are unpaired in the secondary structure;
	\item[\textit{Double}:] Locations where all nucleotides in the corresponding segment are paired in the secondary structure;
	\item[\textit{Transitional}:] Locations where some nucleotides in the corresponding segment are paired and others are unpaired in the secondary structure.
\end{description}
{\em Double} and {\em transitional} locations participate in the candidate secondary structure, while {\em single} locations do not.
As a general rule, local ambiguities measured at any of these locations will increase with the length of the molecule---there are simply more opportunities to find complementary sequences.
Instead of using the local ambiguities themselves, we focus on the {\em differences} between ambiguities in and around stems ({\em double} and {\em transitional} locations) from those at unpaired ({\em single}) locations. In particular, we defined the ``T-S ambiguity index'' to be the difference between the average local ambiguity at {\em transitional} locations minus the average at {\em single} locations. We defined the ``D-S ambiguity index'' similarly, but adjusted for the evident bias at {\em double} locations, each of which, after all, has at least one complementary pair somewhere in the molecule. The D-S ambiguity index, then, is the average local ambiguity at {\em double} locations minus the average over those {\em single} locations that have at least one ambiguity.

Using only primary and (comparative-analysis) secondary structures, elementary counting statistics, 
and exact, distribution-free, tests,
we will give evidence that
both the T-S and D-S ambiguity indexes 
significantly separate two groups of non-coding RNA molecules: one group consists of  RNA families that operate, {\em in vivo}, as single entities---the Group I and Group II Introns;
the other group is made up of RNA families  known to be active as protein-RNA complexes (i.e. as ribonculeoproteins)---the transfer-messenger RNAs (tmRNA), the RNAs of signal recognition particles (SRP RNA), the ribonuclease P family (RNase P), and the 16s and 23s ribosomal RNAs (16s and 23s rRNA). In particular, ``unbound'' RNA molecules, which perform their functions without being part of a larger nucleoprotein complex, have systematically lower T-S and D-S ambiguity indexes than the ``bound'' RNA molecules found in ribonucleoproteins. 
There are many possible explanations---see {\em Discussion}. Possibly, the unbound molecules are more sensitive to ambiguities and their energy traps, especially those ambiguities that involve the structurally critical {\em double} and {\em transitional} regions, then the bound molecules, whose secondary structures may well be influenced by their chemical relationships to proteins. Interestingly, this distinction between the ambiguity indexes of bound and unbound molecules largely disappears when MFE structures are used instead of comparative-analysis structures in defining {\em double}, {\em transitional}, and {\em single} locations. In fact, for most unbound molecules we can classify a candidate secondary
structure (was it derived from comparative analysis or by a minimum-free-energy calculation?) just by looking at the difference in the T-S or the D-S index under the two structures.

>>>>

Because of the time-consuming nature of experimental determination of RNA structures, a considerable amount of work has been put into computational (as opposed to experimental) approaches. For secondary structure prediction, comparative analysis is the gold standard  \cite{Gutell1992-hu}. But accurate comparative analysis typically requires a large number of homologous sequences, which may or may not be available. On the other hand, the prevailing {\em computational} approaches  xxx References???xxx are based on an important and controversial assumption. Namely, that the structures of  non-coding RNA molecules, {\em in vivo}, are in thermal equilibrium. If this were the case, and if the configuration energies could be accurately specified, then secondary (and in principle, even tertiary) structures could be explored through Monte Carlo sampling from the Boltzmann equilibrium distribution. Alternatively, depending on the nature of the energy landscape, secondary structures could be approximated by identifying the minimum free energy (MFE) configuration  \cite{Mathews1999-hc,Zuker1999-rc}. However, even if we put aside the necessity for approximations, the biological relevance of equilibrium configurations has been a source of misgivings at least since 1969, when Levinthal pointed out that the time required to equilibrate might be too long by many orders of magnitude  \cite{Levinthal1969-hi}. In light of these observations, and considering the ``frustrated'' nature of the folding landscape, many have argued that when it comes to structure prediction for macromolecules, kinetic accessibility is more relevant than equilibrium thermodynamics  \cite{Higgs2000-gj,Flamm2008-ir,Baker1994-px}.

@@

We will introduce quantitative measures of ambiguity and demonstrate their statistical relationships to native secondary structure and to qualitative distinctions between RNA families.
Using a somewhat arbitrary convention, we will
refer to sequences of length four (four consecutive nucleotides) as {\em segments}. Keeping this convention in mind, each line of statistical
evidence is based on the same formal definition of the ``local ambiguity'' of a given segment. This is simply the number of its complementary pairs in the molecule. 
When we refer to the {\em location} of a segment, we will mean the location of the first element of the segment, counting from 5$^\prime$ to 3$^\prime$, and  
when we refer to the local ambiguity of a location, we will mean the local ambiguity of the segment at that location. 
Local ambiguity is an intrinsic property of the primary structure. We will be interested in exposing its statistical relationships to any given candidate secondary structure of the same molecule. For this purpose, for any particular secondary structure we distinguish three different kinds of locations:
\begin{description}
	\item[\textit{Single}:] Locations where all nucleotides in the corresponding segment are unpaired in the secondary structure;
	\item[\textit{Double}:] Locations where all nucleotides in the corresponding segment are paired in the secondary structure;
	\item[\textit{Transitional}:] Locations where some nucleotides in the corresponding segment are paired and others are unpaired in the secondary structure.
\end{description}
{\em Double} and {\em transitional} locations participate in the candidate secondary structure, while {\em single} locations do not.
As a general rule, local ambiguities measured at any of these locations will increase with the length of the molecule---there are simply more opportunities to find complementary sequences.
Instead of using the local ambiguities themselves, we focus on the {\em differences} between ambiguities in and around stems ({\em double} and {\em transitional} locations) from those at unpaired ({\em single}) locations. In particular, we defined the ``T-S ambiguity index'' to be the difference between the average local ambiguity at {\em transitional} locations minus the average at {\em single} locations. We defined the ``D-S ambiguity index'' similarly, but adjusted for an evident bias at {\em double} locations, most of which will have at least one complementary pair somewhere in the molecule. 
(The exceptions being bulges, which occur in about 20\% of the stems in our data set.)
With this in mind, we defined the D-S ambiguity index to be the average local ambiguity at {\em double} locations minus the average over those {\em single} locations that have at least one ambiguity\footnote{
An argument could be made for including bulges in the definition of the {\em local ambiguity} in the first place. We have not explored this or other variations, though there is not shortage of arguably useful descriptive statistics for capturing the notion of ambiguity.}.

Using only primary and (comparative-analysis) secondary structures, elementary counting statistics, 
and exact, distribution-free, tests,
we will give evidence that
both the T-S and D-S ambiguity indexes 
significantly separate two groups of non-coding RNA molecules: one group consists of  RNA families that operate, {\em in vivo}, as single entities---the Group I and Group II Introns;
the other group is made up of RNA families  known to be active as protein-RNA complexes (i.e. as ribonculeoproteins)---the transfer-messenger RNAs (tmRNA), the RNAs of signal recognition particles (SRP RNA), the ribonuclease P family (RNase P), and the 16s and 23s ribosomal RNAs (16s and 23s rRNA). In particular, ``unbound'' RNA molecules, which perform their functions without being part of a larger nucleoprotein complex, have systematically lower T-S and D-S ambiguity indexes than the ``bound'' RNA molecules found in ribonucleoproteins. 
There are many possible explanations (see {\em Discussion}). From the kinetic viewpoint, one possibility is that unbound molecules are more sensitive to ambiguities and their energy traps, especially those ambiguities that involve the structurally critical {\em double} and {\em transitional} regions. The secondary structures of bound molecules, on the other hand, are likely influenced by their chemical relationships to proteins. Perhaps, therefore, they are less sensitive to forming unwanted stems. Interestingly, this distinction between the ambiguity indexes of bound and unbound molecules largely disappears when MFE structures are used instead of comparative-analysis structures in defining {\em double}, {\em transitional}, and {\em single} locations. In fact, for most unbound molecules we can classify a candidate secondary
structure (was it derived from comparative analysis or by a minimum-free-energy calculation?) just by looking at the difference in the T-S or the D-S index under the two structures.
